PTFO 8-8695: Two Stars, Two Signals, No Planet

### 2020-05-14 paper/ms.tex
ditto

Performing the same analysis using the renormalized unit weight error
(RUWE\footnote{ See the Gaia DPAC technical note
GAIA-C3-TN-LU-LL-124-01,
\url{http://www.rssd.esa.int/doc_fetch.php?id=3757412}, accessed
2020-04-27. }) rather than the reduced $\chi^2$ yielded similar
results.  \ptfo\ has a RUWE of 1{.}22, which corresponds to the
93$^{\rm rd}$ percentile of 25$\,$Ori-1 members.  Two of thirty stars
with variability amplitudes greater than 9.7\% showed higher RUWE.
One was CVSO~35, which has a TESS light curve that varies by 2
magnitudes, and shows a strong IR excess and a 10$\mu$m silicate
emission feature \citep{mauco_herschel_2018}.  The other is GAIA DR2
3222210363837122048.

@@

We are not sure what to make of the numerical coincidence.  The ratio
of long to short periods is roughly 10:9.  It is not clear that this
would obviously translate into an observational bias unless, by some
fluke, three season's worth of observations managed to only observe
every ninth dip.  This is of course not the case, and we therefore
leave this curiosity as observation {\it sans} interpretation.
% JNW: This seems like an important clue. I don't have the mental
% energy to ponder it now, but it deserves pondering.
% JNW: Could this mean that both stars are dippers? A change in period
% is degenerate with a change in period, over a short interval, right?
% so maybe one set of dips belongs to one star, the other set to the 2nd star?
% Features with a different period would appear to lie along a line
% of constant slope in this diagram...what would the slope be, in this case,
% if the feature period were P_long?

@@

These light curves have been phenomenologically classified as
``persistent flux dips'' or ``transient flux dips'', based on whether
their depths and durations show variability over the 90-day K2
campaigns \citep{stauffer_orbiting_2017}.  In the terminology of
\citet{stauffer_orbiting_2017}, these objects are morphologically
distinct from ``scallop shell'' light curves, and are present in stars
at more advanced evolutionary disk stages than the ``dipper'' stars
\citep{ansdell_young_2016,cody_manyfaceted_2018}.  The persistent and
transient flux dip stars all show angular dips that are cannot be
explained as the effects of starspots.  These stars typically have the
following things in common:
\begin{enumerate}[topsep=0.5ex,itemsep=-0.5ex,partopsep=1ex,parsep=1.5ex]
  \item They are weak-lined T Tauri stars.
  \item The spectral type is M2 to M5 ({\it e.g.},
    \citealt{rebull_usco_2018},~Figure~20).
  \item The age is typically $\lesssim$ 100$\,$Myr.\footnote{
    At present, the oldest observed ``scallops'' are in the Pleaides
    \citep{rebull_rotation_2016}. One of these, EPIC 211013604, might
    meet the ``persistent dip'' classification.  If so, it is also the
    oldest known.} 
  \item The light curves show shallow, angular dips, usually superposed
    on large-amplitude smooth variability. The latter is interpreted
    as stellar rotation.
  \item The rotation is rapid, with a period that is usually between 0.5 and 2.0 days.
  \item There is rarely any infrared excess that is detectable in the WISE data
  (never any W4 detection; only a few W3 detections).
  % JNW: Spell out acronym and give reference?
  \item They sometimes show multiple dips per cycle.
  \item The dip depths, durations, and phases can vary over just a few
    cycles ({\it e.g.}, EPIC 204143627).
  \item The dip depths can change after flares.
  \item They are rare at a population level, with an occurrence rate of $\lesssim 1\%$ of
  the relevant stars \citep{rebull_usco_2018}.
% JNW: what is 'relevant stars'? Can this be replaced by 'young M2-M5 stars'?
\end{enumerate}
The 10.76$\,$hr signal in \ptfo\ meets all of these criteria.  This is
the first connection of \ptfo\ with this class of objects because the
TESS light curve was needed to resolve the different rotation signals.
% JNW: Despite the preceding sentence, I don't understand why this connection was impossible before TESS.

@@

Based on the lack of an infrared excess seen by \citet{yu_tests_2015},
the primordial gas disks seem to be have been depleted\footnote{A
potential confounding factor is completeness. Is Spitzer sensitive to
infrared excesses at the distance of Orion? The series of studies by
\cite{hernandez_spitzer_2006,hernandez_spitzer_ob1_2007,hernandez_spitzer_sig_2007,hernandez_spitzer_2009}
show that it is.}
% JNW: I don't understand why we can be so definitive...
% an infrared excess is not either 'present' or 'absent'. The upper limit in Jy
% corresponds to a constraint on the mass/radius/etc of any dusty disk.
around both stars in \ptfo. The stars are therefore
no longer magnetically locked to their disks.  This is consistent with
the $\approx$half-day periodicities of both rotation signals: young
disked M dwarfs typically rotate with periods of two days or more due
to magnetic locking \citep[{\it e.g.},][]{rebull_rotation_2020}.  If
the two stars are within $\approx$50$\,$AU of each other, as required
by the NIRC2 adaptive optics imaging, then it would also be expected
that the stars would truncate the outer edges of their respective
disks, in a manner seen at the population level in exoplanetary
systems \citep{kraus_impact_2016,moe_impact_2019}.  This hard outer
boundary condition could propagate to the inner disk and affect its
evolution.
% JNW: I don't understand the last sentence.

The key question is what causes the transient dips. This is
an unsolved problem not only for \ptfo\ but also for the emerging class
of similar young and rapidly rotating M-dwarfs.  Many possible
explanations were discussed by \citet{rebull_rotation_2016},
\citet{david_transient_2017}, \citet{stauffer_orbiting_2017}, and
\citet{zhan_complex_2019}.  Among the {\it disfavored} explanations are that
the dips are caused by
{\it (i)} eclipsing binaries;
{\it (ii)} ``dipper''-flavor Class-I or Class-II disks;
{\it (iii)} eclipses of prominences;
% JNW: why is this one ruled out? Sounds similar to the accretion-hotspot hypothesis.
{\it (iv)} high-latitude accretion hotspots;
{\it (v)} high-latitude starspots;
or
{\it (vi)} dust clouds of plausible composition.
We also view the possibility of {\it (vii)} tidally disrupted
planetary or cometary material to be implausible, given the
synchronicity between dip and rotation periods seen across many
systems.
% JNW Could the material get magnetically locked at the corotation radius?

>>>>

Performing the same analysis using the renormalized unit weight error
(RUWE\footnote{ See the Gaia DPAC technical note
GAIA-C3-TN-LU-LL-124-01,
\url{http://www.rssd.esa.int/doc_fetch.php?id=3757412}, accessed
2020-04-27. }) rather than the reduced $\chi^2$ yielded similar
results.  \ptfo\ has a RUWE of 1{.}22, which corresponds to the
93$^{\rm rd}$ percentile of 25$\,$Ori-1 members.  Two of thirty stars
with variability amplitudes greater than 9.7\% showed higher RUWE.
One was CVSO~35, which has a TESS light curve that varies by 2
magnitudes.
CVSO~35 also shows a strong Wide-field Infrared Survey Exoplorer
(WISE, \citealt{wright_WISE_2010}) IR excess as well as a 10$\mu$m silicate
emission feature \citep{mauco_herschel_2018}.  The other is GAIA DR2
3222210363837122048.

@@

We are not sure what to make of the numerical coincidence.  The ratio
of long to short periods is roughly 10:9.  It is not clear that this
would obviously translate into an observational bias unless, by some
fluke, three season's worth of observations managed to only observe
every ninth dip.  This is of course not the case, and we therefore
leave this curiosity as observation {\it sans} interpretation.
% JNW: This seems like an important clue. I don't have the mental
% energy to ponder it now, but it deserves pondering.
% JNW: Could this mean that both stars are dippers? A change in period
% is degenerate with a change in period, over a short interval, right?
% so maybe one set of dips belongs to one star, the other set to the 2nd star?
% Features with a different period would appear to lie along a line
% of constant slope in this diagram...what would the slope be, in this case,
% if the feature period were P_long?
% LGB: Seems unlikely, because we're seeing a something with constant
% 10.74 hour periodicity, but the phase is offset by 1/9th.
% If both stars were dippers, we would be seeing a 11.98 hour
% periodicity instead.

@@

These light curves have been phenomenologically classified as
``persistent flux dips'' or ``transient flux dips'', based on whether
their depths and durations show variability over the 90-day K2
campaigns \citep{stauffer_orbiting_2017}.  In the terminology of
\citet{stauffer_orbiting_2017}, these objects are morphologically
distinct from ``scallop shell'' light curves, and are present in stars
at more advanced evolutionary disk stages than the ``dipper'' stars
\citep{ansdell_young_2016,cody_manyfaceted_2018}.  The persistent and
transient flux dip stars all show angular dips that are cannot be
explained as the effects of starspots.  These stars typically have the
following things in common:
\begin{enumerate}[topsep=0.5ex,itemsep=-0.5ex,partopsep=1ex,parsep=1.5ex]
  \item They are weak-lined T Tauri stars.
  \item The spectral type is M2 to M5 ({\it e.g.},
    \citealt{rebull_usco_2018},~Figure~20).
  \item The age is typically $\lesssim$ 100$\,$Myr.\footnote{
    At present, the oldest observed ``scallops'' are in the Pleaides
    \citep{rebull_rotation_2016}. One of these, EPIC 211013604, might
    meet the ``persistent dip'' classification.  If so, it is also the
    oldest known.} 
  \item The light curves show shallow, angular dips, usually superposed
    on large-amplitude smooth variability. The latter is interpreted
    as stellar rotation.
  \item The rotation is rapid, with a period that is usually between 0.5 and 2.0 days.
  \item There is rarely any infrared excess that is detectable in the WISE data
  (never any W4 detection; only a few W3 detections).
  \item They sometimes show multiple dips per cycle.
  \item The dip depths, durations, and phases can vary over just a few
    cycles ({\it e.g.}, EPIC 204143627).
  \item The dip depths can change after flares.
  \item They are rare at a population level, with an occurrence rate of $\lesssim 1\%$ of
  young M2-M5 stars \citep{rebull_usco_2018}.
\end{enumerate}
The 10.76$\,$hr signal in \ptfo\ meets all of these criteria.  This is
the first connection of \ptfo\ with this class of objects, 
likely because the TESS data enabled us to put the dips in 
the context of the asymmetric out-of-dip modulation.

@@

Based on the lack of an infrared excess seen by \citet{yu_tests_2015},
the primordial gas disks seem to be have been depleted\footnote{A
potential confounding factor is completeness. Is Spitzer sensitive to
infrared excesses at the distance of Orion? The series of studies by
\cite{hernandez_spitzer_2006,hernandez_spitzer_ob1_2007,hernandez_spitzer_sig_2007,hernandez_spitzer_2009}
show that it is.}
% JNW: I don't understand why we can be so definitive...
% an infrared excess is not either 'present' or 'absent'. The upper limit in Jy
% corresponds to a constraint on the mass/radius/etc of any dusty disk.
around both stars in \ptfo. The stars are therefore no longer
magnetically locked to their disks.  This is consistent with the
$\approx$half-day periodicities of both rotation signals: young disked
M dwarfs typically rotate with periods of two days or more due to
magnetic locking \citep[{\it e.g.},][]{rebull_rotation_2020}.  If the
two stars are within $\approx$50$\,$AU of each other, as required by
the NIRC2 adaptive optics imaging, then it would also be expected that
the stars would have truncated the outer edges of their respective
disks, in a manner seen at the population level in exoplanetary
systems \citep{kraus_impact_2016,moe_impact_2019}.

The key remaining question is what causes the transient dips. This is
an unsolved problem not only for \ptfo\ but also for the emerging
class of similar young and rapidly rotating M-dwarfs.  Many possible
explanations were discussed by \citet{rebull_rotation_2016},
\citet{david_transient_2017}, \citet{stauffer_orbiting_2017}, and
\citet{zhan_complex_2019}.  Among the {\it disfavored} explanations
are that the dips are caused by
{\it (i)} eclipsing binaries;
{\it (ii)} ``dipper''-flavor Class-I or Class-II disks;
{\it (iii)} eclipses of prominences;
% JNW: why is this one ruled out? Sounds similar to the accretion-hotspot hypothesis.
{\it (iv)} high-latitude accretion hotspots;
{\it (v)} high-latitude starspots;
or
{\it (vi)} dust clouds of plausible composition.
We also view the possibility of {\it (vii)} tidally disrupted
planetary or cometary material to be implausible, given the
synchronicity between dip and rotation periods seen across many
systems.
% JNW Could the material get magnetically locked at the corotation radius?
